A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-28 text/manuscript.Rmd
fix textual figure references; reflow text

We investigated this dataset further using multiple correspondence analysis
(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
2** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. 
We used the resulting UMAP graph coordinates to perform
density based clustering with HDBSCAN [@hdbscan], which resulted in
identification of 10 well-defined clusters capturing about 80% of the tumor
samples. To capture the remaining samples into one of these 10 clusters we used
kNN (see **Supplementary Methods** for details on clustering methods).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregate into one or more predominant
clusters. To our surprise, most cancer types were not heavily biased in one
cluster, and all well-represented cancer types had tumors in every cluster (see
**figure 3A** and full tumor profiles in **supplementary figure S1**, see also
interactive media from **supplementary file 1**), suggesting that our
pathway-disruptions identify clusters of molecular pathology largely independent
of tissue-of-origin. As an example of one type of cancer that does have a biased
pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
cluster 8 (**figure 3A** and **supplementary figure S1** and **supplementary
file 1**). But even PAAD comprises tumors from the nine remaining classes,
meaning that patients with these tumors have potentially different underlying
molecular pathologies.

@@

#### Figure 3
```{r child = '../code/figure3.Rmd'}
```

**Figure 3: Pathway-based clustering independent of tissue-of-origin** 
A) Sankey plot showing correspondence between cancer type and cluster identity.
B) Projection of breast cancer subtypes onto the UMAP. C) Projection of BRCA1/2
somatic mutation onto the UMAP. D) Projection of tumor stage onto the UMAP,
regardless of cancer type. E) Projection of metastatic status onto the UMAP.
Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors. F) Somatic mutation 
frequencies for each cluster. Vertical axis shows log mutation count, horizontal 
axis is cluster identity. Each dot represents an individual tumor sample, ranked 
lowest to highest by mutation count. The median mutation count in each cluster is 
indicated by the horizontal line.

@@

There are several well-known familial cancer-causing mutations that have been
studied for differences in basic biology, survival and treatment outcomes. The
functions of these genes are related to risk factors such as genome stability,
proof-reading & DNA damage repair, and telomere length. _BRCA1/2_ genes for
example are key for DNA double-stranded break repair
[@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in these genes
confer elevated risk for breast, prostate and ovarian cancers. The mechanism of
risk is thought to involve loss of heterozygosity resulting in loss of the
wildtype, functional allele [@brca-loh], so we projected the somatic mutations
for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe segregation of
these mutations into specific clusters (**figure 3C** and see interactive media
from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar
projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_,
_PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes
except for _APC_ exhibited any remarkable specificity with respect to cluster
assignment (**supplementary figure S3**). To look at other risk factors such as
maintenance of DNA methylation levels and telomere length, we projected somatic
mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and
observed similar lack of segregation by cluster (**supplementary figure S3**).

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these criteria, it is possible that more advanced tumors have
common pathway disruption profiles. The UMAP, which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle reflect progression through a series of stages. The backbone starts
with a cluster of tumors (class 1) that has the fewest point mutations and
culminates in a cluster (class 10) which has nearly every pathway disrupted
(**figure 4**). However, outside of class 10 we don't observe a trend in the
overall mutation burden across the backbone of the UMAP. Nonetheless, to test
the hypothesis that the molecular-pathway disruption clusters represent
advancement through stages, we projected staging data onto the UMAP. Similar to
tissue of origin and other categories of tumor, we did not observe any bias
among the stages to specific clusters (**figure 3D**), suggesting that stage is
not a contributing factor to cluster identity.

Finally, as a measure of tumor advancement, metastasis is the condition in which
certain phenotypic criteria are met: loss of
differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune
system evasion and tissue invasiveness [@hallmarks_II]. To determine whether any
of our clusters correspond to an especially advanced stage of cancer across
tissue types, we projected the metastases onto our UMAP, and surprisingly
we observed even distribution of the samples across clusters (**figure 3E**).
This final observation suggests that our pathway-disruption clustering is
dependent on particular combinations of gene mutations affecting different
pathways that can each give rise to advanced stages of disease and metastasis,
regardless and independent of overall mutational burden.

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment _within cluster_ relative to all other clusters (see methods)
and ranked pathways from highest to lowest enrichment. We visualized the
enrichment as a heatmap (**figure 5A**). Using this approach, we identified
about fourteen pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see
methods) (**supplementary table ST2**). Clusters 7, 8, and 9 had several
pathways in common. To explore the specific pathways marking each cluster, we
projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4** and see interactive media from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Clusters 3 and 5
were distinguished by metabolic pathways including RNA and protein biosynthesis
(**supplementary figure S4**). Similarly, cluster 4 was distinguished by
mutations affecting regulation of DNA and histone methylation ("DNA
methylation", "PRC2 methylates histones and DNA", and "Nucleosome assembly").
Clusters 7-9 have in common mutations in extracellular, intracellular, and
immune-related signaling pathways (see **figure 5B** and **supplementary figure
S4**). Cluster 2 had the highest pathway enrichment levels of the three, having
mutations in hedgehog signaling, "β-catenin degradation", "cellular response to
hypoxia", "regulation of cell cycle" and "apoptosis" among others.

@@

#### Figure 5
```{r child = '../code/figure5.Rmd'}
```

**Figure 5: Pan-cancer enrichment of pathway disruptions.** 
A) Heatmap shows the relative enrichment of each pathway (rows) within each
numbered cluster (columns). Effect size is displayed as colors representing
percent enrichment. B) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size ≥ 0.30 in each
cluster) within each numbered cluster (rows).

@@

Prior efforts to extract signatures from pan-cancer datasets met with difficulty
in distinguishing tumor samples from their tissue-specific -omics data
signatures. Given our pathway-disruption based clustering, this raises the
question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (_i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5** and see interactive media from
[**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
cluster 9, which also shared multiple enrichment in signaling pathways with
clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5,
defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
and nuclear pore complex, respectively. Cluster 4 had mutations in genes
responsible for nucleosome structure. Cluster 2 had mutations in proteasomal
subunit genes involved in protein degradation. We also observed that genes that
were enriched for one cluster are depleted from others (_i.e._ is enriched in
cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but
depleted in clusters 3 and 8). Next, we investigated the proportion of samples
per cancer type for the significant genes within a cluster (**Figure 6**;
heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
more highly mutated genes across all cancers. Instead, when observing the
mutation rate for these genes within samples that belong to a cluster, the
mutation rate is heterogeneous across tumors by tissue origin (_e.g._ In cluster
4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was
enriched for _HIST1H2AC_). Even among the top enriched genes within clusters
there is no global pattern, indicating that our clusters are not driven by
individual genes, but rather networks as a whole. Taken together, our data
identify a framework of cancer type-specific mutations associated with specific
clusters.

@@

#### Figure 6
```{r child = '../code/figure6.Rmd'}
```

**Figure 6: Gene level analysis reveals tissue-specific class signatures.** 
Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
95% confidence interval. Rows: significant genes from each cluster. Each cluster
was compared against the background (all other clusters) to find differentially
mutated genes. Significant genes (pvalue < 0.01) were selected and limited to
the top ten results for each cluster. Heatmap; Columns: cancer type. Rows:
significant genes. The heatmap shows proportion of samples per cluster and
cancer type mutated for each gene. Depleted significant genes (left). Enriched
significant genes (right).

>>>>

We investigated this dataset further using multiple correspondence analysis
(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
1B** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. 
We used the resulting UMAP graph coordinates to perform
density based clustering with HDBSCAN [@hdbscan], which resulted in
identification of 10 well-defined clusters capturing about 80% of the tumor
samples. To capture the remaining samples into one of these 10 clusters we used
kNN (see **Supplementary Methods** for details on clustering methods).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregate into one or more predominant
clusters. To our surprise, most cancer types were not heavily biased in one
cluster, and all well-represented cancer types had tumors in every cluster (see
**figure 2A** and full tumor profiles in **supplementary figure S1**, see also
interactive media from **supplementary file 1**), suggesting that our
pathway-disruptions identify clusters of molecular pathology largely independent
of tissue-of-origin. As an example of one type of cancer that does have a biased
pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
cluster 8 (**figure 2A** and **supplementary figure S1** and **supplementary
file 1**). But even PAAD comprises tumors from the nine remaining classes,
meaning that patients with these tumors have potentially different underlying
molecular pathologies.

@@

#### Figure 2
```{r child = '../code/figure3.Rmd'}
```

**Figure 2: Pathway-based clustering independent of tissue-of-origin** 
A) Sankey plot showing correspondence between cancer type and cluster identity.
B) Projection of breast cancer subtypes onto the UMAP. C) Projection of BRCA1/2
somatic mutation onto the UMAP. D) Projection of tumor stage onto the UMAP,
regardless of cancer type. E) Projection of metastatic status onto the UMAP.
Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors. F) Somatic mutation 
frequencies for each cluster. Vertical axis shows log mutation count, horizontal 
axis is cluster identity. Each dot represents an individual tumor sample, ranked 
lowest to highest by mutation count. The median mutation count in each cluster is 
indicated by the horizontal line.

@@

There are several well-known familial cancer-causing mutations that have been
studied for differences in basic biology, survival and treatment outcomes. The
functions of these genes are related to risk factors such as genome stability,
proof-reading & DNA damage repair, and telomere length. _BRCA1/2_ genes for
example are key for DNA double-stranded break repair
[@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in these genes
confer elevated risk for breast, prostate and ovarian cancers. The mechanism of
risk is thought to involve loss of heterozygosity resulting in loss of the
wildtype, functional allele [@brca-loh], so we projected the somatic mutations
for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe segregation of
these mutations into specific clusters (**figure 2C** and see interactive media
from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar
projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_,
_PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes
except for _APC_ exhibited any remarkable specificity with respect to cluster
assignment (**supplementary figure S3**). To look at other risk factors such as
maintenance of DNA methylation levels and telomere length, we projected somatic
mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and
observed similar lack of segregation by cluster (**supplementary figure S3**).

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these criteria, it is possible that more advanced tumors have
common pathway disruption profiles. The UMAP, which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 1B**) could
in principle reflect progression through a series of stages. The backbone starts
with a cluster of tumors (class 1) that has the fewest point mutations and
culminates in a cluster (class 10) which has nearly every pathway disrupted
(**figure 2F**). However, outside of class 10 we don't observe a trend in the
overall mutation burden across the backbone of the UMAP. Nonetheless, to test
the hypothesis that the molecular-pathway disruption clusters represent
advancement through stages, we projected staging data onto the UMAP. Similar to
tissue of origin and other categories of tumor, we did not observe any bias
among the stages to specific clusters (**figure 2D**), suggesting that stage is
not a contributing factor to cluster identity.

Finally, as a measure of tumor advancement, metastasis is the condition in which
certain phenotypic criteria are met: loss of
differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune
system evasion and tissue invasiveness [@hallmarks_II]. To determine whether any
of our clusters correspond to an especially advanced stage of cancer across
tissue types, we projected the metastases onto our UMAP, and surprisingly
we observed even distribution of the samples across clusters (**figure 2E**).
This final observation suggests that our pathway-disruption clustering is
dependent on particular combinations of gene mutations affecting different
pathways that can each give rise to advanced stages of disease and metastasis,
regardless and independent of overall mutational burden.

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment _within cluster_ relative to all other clusters (see methods)
and ranked pathways from highest to lowest enrichment. We visualized the
enrichment as a heatmap (**figure 3A**). Using this approach, we identified
about fourteen pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see
methods) (**supplementary table ST2**). Clusters 7, 8, and 9 had several
pathways in common. To explore the specific pathways marking each cluster, we
projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4** and see interactive media from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Clusters 3 and 5
were distinguished by metabolic pathways including RNA and protein biosynthesis
(**supplementary figure S4**). Similarly, cluster 4 was distinguished by
mutations affecting regulation of DNA and histone methylation ("DNA
methylation", "PRC2 methylates histones and DNA", and "Nucleosome assembly").
Clusters 7-9 have in common mutations in extracellular, intracellular, and
immune-related signaling pathways (see **figure 3B** and **supplementary figure
S4**). Cluster 2 had the highest pathway enrichment levels of the three, having
mutations in hedgehog signaling, "β-catenin degradation", "cellular response to
hypoxia", "regulation of cell cycle" and "apoptosis" among others.

@@

#### Figure 3
```{r child = '../code/figure5.Rmd'}
```

**Figure 3: Pan-cancer enrichment of pathway disruptions.** 
A) Heatmap shows the relative enrichment of each pathway (rows) within each
numbered cluster (columns). Effect size is displayed as colors representing
percent enrichment. B) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size ≥ 0.30 in each
cluster) within each numbered cluster (rows).

@@

Prior efforts to extract signatures from pan-cancer datasets met with difficulty
in distinguishing tumor samples from their tissue-specific -omics data
signatures. Given our pathway-disruption based clustering, this raises the
question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (_i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5** and see interactive media from
[**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
cluster 9, which also shared multiple enrichment in signaling pathways with
clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5,
defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
and nuclear pore complex, respectively. Cluster 4 had mutations in genes
responsible for nucleosome structure. Cluster 2 had mutations in proteasomal
subunit genes involved in protein degradation. We also observed that genes that
were enriched for one cluster are depleted from others (_i.e._ is enriched in
cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but
depleted in clusters 3 and 8). Next, we investigated the proportion of samples
per cancer type for the significant genes within a cluster (**Figure 4**;
heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
more highly mutated genes across all cancers. Instead, when observing the
mutation rate for these genes within samples that belong to a cluster, the
mutation rate is heterogeneous across tumors by tissue origin (_e.g._ In cluster
4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was
enriched for _HIST1H2AC_). Even among the top enriched genes within clusters
there is no global pattern, indicating that our clusters are not driven by
individual genes, but rather networks as a whole. Taken together, our data
identify a framework of cancer type-specific mutations associated with specific
clusters.

@@

#### Figure 4
```{r child = '../code/figure6.Rmd'}
```

**Figure 4: Gene level analysis reveals tissue-specific class signatures.** 
Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
95% confidence interval. Rows: significant genes from each cluster. Each cluster
was compared against the background (all other clusters) to find differentially
mutated genes. Significant genes (pvalue < 0.01) were selected and limited to
the top ten results for each cluster. Heatmap; Columns: cancer type. Rows:
significant genes. The heatmap shows proportion of samples per cluster and
cancer type mutated for each gene. Depleted significant genes (left). Enriched
significant genes (right).

### 2020-08-28 text/survival.Rmd
fix textual figure references; reflow text

To illustrate the model, in **Figure 7A** we simulate survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

#### Figure 7
```{r child='../code/survival/survfig.Rmd'}
```

**Figure 7: Comparison of different cancers, simulated and actual results
results.** To illustrate the risk model, we compare survival of four simulated
cohorts. A) A group consisting of randomly selected 30 year old women (light
blue) vs. an equal sized group of 30 year old women with cancer, k = 2 (light
orange). To illustrate how the mix of ages alters the survival curve, we compare
a randomly selected group of people with ages uniformly distributed from 30 to
70 (dark blue) and a randomly selected group with uniformly distributed ages and
cancer k = 2 (dark orange). B,C,D,E) We compare a randomly selected group with
correct age distribution for the given cancer (turqoise) to the model predicted
survival for the given average k (blue) and the actual survival for the patients
in our dataset (orange). Differences with actual data are likely due to a
mixture of cluster identities which we exclude from these simulations for
simplicity. F) Effects are compared between cancer type by
showing the probability density for the posterior value of the $k_{tis}$
multiplier. Larger values correspond to decreased age-independent life
expectancy.

@@

We found that cancer types, as expected, have a range of prognoses relative to
the general population. In **Figure 7B-E** we see three particularly deadly
cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
where diagnosis apparently decreases risk relative to background (Thyroid:
THCA). Cancers with posterior probability for relative risk of less than 1 such
as THCA should be interpreted carefully. This Bayesian model is a model for a
state of information. The information that a person is diagnosed with cancer may
lead us to expect that they will live a shorter time than the general population
of matched age (k > 1), or a longer time than the general population of matched
age (k < 1). A shorter prognosis could result from the cancer being aggressive,
injuring the body and causing death. By contrast, a longer prognosis may result
from the cancer being relatively mild, and therefore diagnosis could be an
informational signal that the patient is health conscious, with the comparison
group having more people whose cancers go undiagnosed. It is important to note
therefore that the diagnosis can increase our expectation of life relative to
the comparison group, even if it decreases the expectation of life of the
individual relative to the counterfactual where they did not have cancer.

The degree to which a given cancer accelerates aging can be determined by
the $k_{tis}$ multiplier (**Figure 8**) multiplied by the cluster modifier $k_{cl}$
(**Figure 9**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude.

@@

#### Figure 9

@@

**Figure 9 Cluster specific $k_{cl}$ values for each cancer type.** The overall
$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific value
represents the relative aggressiveness of each cluster within the cancer type.
For example clusters 1, 2, and 6 Glioblastoma (GBM) are apparently much less
deadly than other types of GBM. A similar situation can be seen with Pancreatic
Cancer (PAAD) where 1, 6, and 8 are less deadly compared to clusters 7, 9, 10.
For some cancer types there are few data points and little ability to estimate
values with precision (DLBC and KICH for example).

@@

Our estimates of tissue-specific cluster effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
or more deadly than other clusters within each cancer type (**Figure 9**). A
cluster-specific rate of 1 represents the typical rate for this tissue type. For
several cancers (_e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe cluster-specific trends that held
true across cancer types, which could result from different cancers having
different standards of care for example. In support of this interpretation, we
also tested a factored model which considered both cancer type and each cluster
independent of cancer type (not shown). Though we were able to successfully fit
this model, it is a special case of the more general model where cluster and
tissue independently affect longevity, and there is no reason to believe that
cluster specific effects would necessarily remain constant across tissue types
given how widely the patients vary across tissue types, drug and surgical
treatments.

>>>>

To illustrate the model, in **Figure 5A** we simulate survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

#### Figure 5
```{r child='../code/survival/survfig.Rmd'}
```

**Figure 5: Comparison of different cancers, simulated and actual results
results.** To illustrate the risk model, we compare survival of four simulated
cohorts. A) A group consisting of randomly selected 30 year old women (light
blue) vs. an equal sized group of 30 year old women with cancer, k = 2 (light
orange). To illustrate how the mix of ages alters the survival curve, we compare
a randomly selected group of people with ages uniformly distributed from 30 to
70 (dark blue) and a randomly selected group with uniformly distributed ages and
cancer k = 2 (dark orange). B,C,D,E) We compare a randomly selected group with
correct age distribution for the given cancer (turqoise) to the model predicted
survival for the given average k (blue) and the actual survival for the patients
in our dataset (orange). Differences with actual data are likely due to a
mixture of cluster identities which we exclude from these simulations for
simplicity. F) Effects are compared between cancer type by showing the
probability density for the posterior value of the $k_{tis}$ multiplier. Larger
values correspond to decreased age-independent life expectancy.

@@

We found that cancer types, as expected, have a range of prognoses relative to
the general population. In **Figure 5B-E** we see three particularly deadly
cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
where diagnosis apparently decreases risk relative to background (Thyroid:
THCA). Cancers with posterior probability for relative risk of less than 1 such
as THCA should be interpreted carefully. This Bayesian model is a model for a
state of information. The information that a person is diagnosed with cancer may
lead us to expect that they will live a shorter time than the general population
of matched age (k > 1), or a longer time than the general population of matched
age (k < 1). A shorter prognosis could result from the cancer being aggressive,
injuring the body and causing death. By contrast, a longer prognosis may result
from the cancer being relatively mild, and therefore diagnosis could be an
informational signal that the patient is health conscious, with the comparison
group having more people whose cancers go undiagnosed. It is important to note
therefore that the diagnosis can increase our expectation of life relative to
the comparison group, even if it decreases the expectation of life of the
individual relative to the counterfactual where they did not have cancer.

The degree to which a given cancer accelerates aging can be determined by the
$k_{tis}$ multiplier (**Figure 5F**) multiplied by the cluster modifier $k_{cl}$
(**Figure 6**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude.

@@

#### Figure 6

@@

**Figure 6 Cluster specific $k_{cl}$ values for each cancer type.** The overall
$k$ value is the product of $k_{tis} \times k_{cl}$. The cluster-specific value
represents the relative aggressiveness of each cluster within the cancer type.
For example clusters 1, 2, and 6 Glioblastoma (GBM) are apparently much less
deadly than other types of GBM. A similar situation can be seen with Pancreatic
Cancer (PAAD) where 1, 6, and 8 are less deadly compared to clusters 7, 9, 10.
For some cancer types there are few data points and little ability to estimate
values with precision (DLBC and KICH for example).

@@

Our estimates of tissue-specific cluster effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
or more deadly than other clusters within each cancer type (**Figure 6**). A
cluster-specific rate of 1 represents the typical rate for this tissue type. For
several cancers (_e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe cluster-specific trends that held
true across cancer types, which could result from different cancers having
different standards of care for example. In support of this interpretation, we
also tested a factored model which considered both cancer type and each cluster
independent of cancer type (not shown). Though we were able to successfully fit
this model, it is a special case of the more general model where cluster and
tissue independently affect longevity, and there is no reason to believe that
cluster specific effects would necessarily remain constant across tissue types
given how widely the patients vary across tissue types, drug and surgical
treatments.
